Quantum H theorem, variational entropy, and relaxation processes

### 2021-01-26 tex/introduction.tex
Verstion submitted to Entropy on 20210126

The theoretical basis and the procedures that allow to describe equilibrium systems
are well-stablished. These procedures can be applied to treat a wide range of the systems
present in nature, and include both the macroscopic phenomenological methods (thermodynamics)
as well as the microscopic description (statistical mechanics).
(Of course, regarding out-of equilibrium systems, the present situation is still a challenge.) 
For instance,
in the Kinetic Theory of gases, the behavior of a dilute classical gas
is described through the Boltzmann transport equation \cite{bib:huang},
and the time-evolution of a system towards the equilibrium is proved by
means of the Boltzmann $H$-theorem, wherein the
spatial-homogeneous distribution function hypothesis plays a crucial role.

On the other hand, for quantum equilibrium systems, the construction of a framework,
with the same level of
success and universality as the classical version, is still an open problem. For instance,
in order to obtain a complete correspondence principle, between classical mechanics
and quantum mechanics, the quantum analogous of both the Boltzmann $H$-theorem
and the Boltzmann transport equation have not attained a fully accepted form.
In this context, Tolman was one of the earliest physicists to propose a quantum
$H$-theorem \cite{bib:tolman}, using probability transition relations, the random
phases hypothesis, and a spatially homogeneous distribution function.
Also, Tolman proposed a potential quantum analogous of the transport equation,
in terms of the occupation number,
by applying time perturbation theory. Further attempts, under the quantum operator
formalism, have addressed the description of quantum transport phenomena by means of the Hamiltonian
of the system and the master equation (which here is the analogous
of the Boltzmann transport equation)
\cite{bib:grabert1974,bib:wang2014,%
bib:angel2017,
bib:amato2020,bib:nicacio2015,bib:hussein2014}. However, these approaches 
are not consistent with the classical-quantum correspondence principle.
In the same line of thought, some authors have proposed $H$-functionals
and attempted to proof a
quantum $H$-theorem \cite{bib:gorban2014,bib:bennaim2017,bib:silva2010,%
bib:deroeck2006,bib:acharya2019,bib:kastner2017,bib:han2015,bib:das2018,bib:vonneumann2010}.
However, whether or not the homogeneous distribution function hypothesis is assumed, or
if their framework obeys the correspondence principle is unclear or not discussed whatsoever.
Furthermore, the general validity
of the quantum $H$-theorem, the second law of thermodynamics, and the
interpretation of the quantum entropy is still in discussion %[19-26].
\cite{bib:silva2010,bib:acharya2019,bib:kastner2017,bib:han2015,bib:brown2008,%
bib:dragoljub2009,bib:vonneumann2010,%
bib:syros1999,bib:lesovik2016,
bib:lesovik2019}.

On the other hand, the framework to describe spatially non-homogeneous systems is still
under construction, although several approaches have been developed. For instance,
the celebrated Onsager formulation (\textit{aka} linear thermodynamics)
\cite{bib:keizer1987,bib:onsager1931} have been successful in describing irreversible chemical 
and physical phenomena. Nonetheless, some aspects such as describing the internal behavior of glasses
\cite{bib:zanotto2018} and the entropy measurement
\cite{bib:schmelzer2018,bib:nemilov2018} cannot be completely addressed with linear thermodynamics.

@@

In order to contribute to the construction of a consistent classical and
quantum $H$-theorem, within a formalism that describes out-of-equilibrium
non-homogeneous systems, in this article we propose a new theoretical framework.
Specifically, for both classical and quantum systems, we include non-homogeneous distribution functions in
the $H$-functionals, and consider non-homogeneous systems in the proofs of the resulting
$H$-theorems. Our proposed $H$-functionals obey
the correspondence principle, but more importantly, these functionals describe
the time-evolution of spatially non-homogeneous
systems towards equilibrium.

The organization of this article is as follows.
In Section~\ref{sec:classical}, we highlight, to our purposes, the most
fundamental assumptions required to proof the Boltzmann $H$-theorem,
we provide an alternative method to obtain the Maxwell-Boltzmann
distribution using the variational method, followed by our proposal of
an alternative $H$-functional for classical systems, and the demonstration of
the respective $H$-theorem . In Section~\ref{sec:quantum}
we review the Tolman proposal for the quantum
version of the $H$-theorem (quantum $H$-theorem)
and how the Bose-Einstein and Fermi-Dirac distributions are treated within
this framework, subsequently we present our proposal for a quantum $H$-functional
and the proof of the corresponding quantum $H$-theorem. In 
Section~\ref{sec:qccorrespondence}
the classical-quantum correspondence between the
quantum and classical $H$-functionals is analyzed.
In Section~\ref{sec:relaxproc} we explore how relaxation processes occur
in a quantum ideal gas; \commb{here we provide a time-evolution equation that could be
considered the analogous of the classical Boltzmann transport equation.} Finally,
we discuss some key ideas derived from our approach and close with a brief summary
in Section~\ref{sec:disscussion}.

>>>>

The theoretical basis and the procedures that allow to describe equilibrium systems
are well-stablished. These procedures can be applied to treat a wide range of the systems
present in nature, and include both the macroscopic phenomenological methods (thermodynamics)
as well as the microscopic description (statistical mechanics).
(Of course, regarding out-of equilibrium systems, the present situation is still a challenge.)
For instance,
in the Kinetic Theory of gases, the behavior of a dilute classical gas
is described through the Boltzmann transport equation \cite{bib:huang},
and the time-evolution of a system towards the equilibrium is finely accounted for
through the Boltzmann $H$-theorem.

On the other hand, for quantum out-of-equilibrium systems, the construction of a
kinetic framework, with the same level of
success and universality as the classical version, still presents some fundamental challenges. For instance,
in order to obtain a complete correspondence principle, between classical mechanics
and quantum mechanics, the quantum analogous of both the Boltzmann $H$-theorem
and the Boltzmann transport equation have not attained a fully accepted form.
In this context, Tolman was one of the earliest physicists to propose a quantum
$H$-theorem \cite{bib:tolman}, using probability transition relations, the random
phases hypothesis, and an $H$-functional defined in terms of a spatially homogeneous distribution function.
Also, Tolman proposed a potential quantum analogous of the transport equation,
in terms of the occupation numbers,
by applying time perturbation theory. Further attempts, under the quantum operator
formalism, have addressed the description of quantum transport phenomena by means of the Hamiltonian
of the system and the master equation (which is, in these works, the analogous
of the Boltzmann transport equation)
\cite{bib:grabert1974,bib:wang2014,%
bib:angel2017,
bib:amato2020,bib:nicacio2015,bib:hussein2014}. However, these approaches 
are not consistent with the classical-quantum correspondence principle.
In the same line of thought, some authors have proposed $H$-functionals
and attempted to proof a
quantum $H$-theorem \cite{bib:gorban2014,bib:bennaim2017,bib:silva2010,%
bib:deroeck2006,bib:acharya2019,bib:kastner2017,bib:han2015,bib:das2018,bib:vonneumann2010}.
However, whether or not the homogeneous distribution function hypothesis is assumed, or
if their framework fulfills the correspondence principle is unclear or not discussed whatsoever.
After the pioneering work of Tolman, at several stages, there has been some discussion regarding
the general validity of the quantum $H$-theorem,
some possible violations of the second law of thermodynamics, and the
interpretation of the quantum entropy  %[19-26].
\cite{bib:silva2010,bib:acharya2019,bib:kastner2017,bib:han2015,bib:brown2008,%
bib:dragoljub2009,bib:vonneumann2010,%
bib:syros1999,bib:lesovik2016,
bib:lesovik2019}.

On the other hand, the framework to describe spatially non-homogeneous systems is still
under construction, although several approaches have been developed. For instance,
the celebrated Onsager formulation (linear thermodynamics)
\cite{bib:keizer1987,bib:onsager1931} have been successful in describing irreversible chemical 
and physical phenomena. Nonetheless, some topics, such as describing the internal behavior of glasses
\cite{bib:zanotto2018} and the entropy measurement
\cite{bib:schmelzer2018,bib:nemilov2018}, cannot be completely addressed with linear thermodynamics.

@@

In order to contribute to the construction of a consistent classical and
quantum $H$-theorem, within a formalism that describes out-of-equilibrium
non-homogeneous systems, in this article we propose a new theoretical framework.
Specifically, for both classical and quantum systems, we include non-homogeneous distribution functions in
the $H$-functionals, and consider non-homogeneous systems in the proofs of the resulting
$H$-theorems. Our proposed $H$-functionals satisfy
the correspondence principle, but more importantly, these functionals describe
the time-evolution of spatially non-homogeneous
systems towards equilibrium.

The organization of this article is as follows.
In Section~\ref{sec:classical}, we highlight, to our purposes, the most
fundamental assumptions required to proof the Boltzmann $H$-theorem,
we provide an alternative method to obtain the Maxwell-Boltzmann
distribution using the variational method, followed by our proposal of
an alternative $H$-functional for classical systems, and the demonstration of
the respective $H$-theorem . In Section~\ref{sec:quantum}
we review the Tolman proposal for the quantum
version of the $H$-theorem (quantum $H$-theorem)
and how the Bose-Einstein and Fermi-Dirac distributions are treated within
this framework, subsequently we present our proposal for a quantum $H$-functional,
and the proof of the corresponding quantum $H$-theorem. In 
Section~\ref{sec:qccorrespondence}
the classical-quantum correspondence between the
quantum and classical $H$-functionals is analyzed.
In Section~\ref{sec:relaxproc} we explore how relaxation processes occur
in a quantum ideal gas and, on the basis of what we call variational entropy,
we propose a time-evolution equation for the distribution function. Finally,
we discuss some key ideas derived from our approach and close with a brief summary
in Section~\ref{sec:disscussion}.

### 2021-01-26 tex/quantum.tex
Verstion submitted to Entropy on 20210126

Consider a diluted gas of $N$ non-interacting quantum particles (either bosons or fermions),
contained by a vessel of volume
$V$, at some temperature $T$, and total energy E. Starting from the Boltzmann definition
of entropy, the quantum $H$ functional is
%
\begin{equation}
   H_T=-\ln G,
\end{equation}
%
where $G$ describes the total number of accessible quantum states of the gas that satisfy the
above conditions \cite{bib:tolman}.
The quantum $H$-theorem can be demostrated as follows.  $G$ can be divided into groups of neighboring states,
$g_k$, and certain occupation numbers, $n_k$, can be associated with each of these
groups. Thus, the above functional takes the form:
%
\begin{equation}\label{eq:quantumh}
    H_T=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, 
\end{equation}
%
where the upper and lower signs are for bosons and fermions, respectively.
Thus the time derivative of Eq.~(\ref{eq:quantumh}) is
%
\begin{equation}\label{eq:quantumdHdt}
\frac{dH_T}{dt}=\sum_{\kappa}\left[\ln n_{\kappa}-\ln\left(g_{\kappa}\pm n_{\kappa}\right)\right]
\frac{dn_{\kappa}}{dt}.
\end{equation}
%

@@

Consider a diluted gas enclosed by a perfectly isolated vessel of volume $V$, with
total energy $E$, and total number of quantum particles $N$, which can be free fermions or
bosons. For our purposes, the volume $V$ is divided into $K$ small cells, each of which
has a constant volume $\delta V_M=V/K$ ($M=1,\,\dots\,,K$), a temperature $T_M$, an energy $\epsilon_M$,
a number of particles $\mathcal{N}_M$, and a distribution function,
$\{f_{Mn}(t)\}$. Hereafter we will use a short-hand notation:
%
\begin{equation}
   f_{Mn}(t)\equiv f(\vec r_M,\epsilon_{n},t),
\end{equation}
%
where $\vec r_M$ is the radius vector pointing at the center of the $M$-th cell.
$f_{Mn}(t)$ represent the number
of particles contained in the $M$-th cell that occupy the energy level $\epsilon_n$ at
time $t$. Since the particles are considered to be free, the energy levels should not
depend on the cell properties, \textit{i.e.} the energy spectrum, $\{\epsilon_n\}$, is the same for all
cells; thus there is no need to label $\epsilon_n$ with an index $M$.

We propose the following functional as an alternative $H$-functional
for quantum non-homogeneous diluted gases:
%
\begin{eqnarray}\label{eq:qHdef}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\bigg.\nonumber \\
    &&\qquad\qquad\bigg.\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M.
\end{eqnarray}
%
Here, the upper and lower sign refer to fermions and bosons respectively.
In addition, when needed, each cell will have an associated local chemical potential,
$\alpha_M$, and a local $H$-functional, which is defined by:
%
\begin{equation}\label{eq:qHMdef}
   \mathcal{H}_M(t)=\sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)
   \pm \Big(1 \mp f_{Mn}(t)\Big) \ln \Big(1 \mp f_{Mn}(t)\Big) \bigg] \de V_M.
\end{equation}
%
Therefore, $\mathcal{N}_M$ and $\mathcal{E}_M$ as functions of time are given by:
%
\begin{subequations}
\begin{equation}
    {\mathcal{N}}_M(t) = \sum_{n}f_{Mn}(t) \delta V_M
\end{equation}
%
and
%
\begin{equation}
\mathcal{E}_M(t) = \sum_{n}f_{Mn}(t)\epsilon_{n} \de V_M,
\end{equation}
\end{subequations}
%
which are, for the whole system, constrained by the microcanonical restrictions
%
\begin{subequations}\label{eq:qGlobRest}
\begin{equation}\label{eq:qGlobRestN}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\right]\delta  V_M
    =\sum_{M=1}^{K}\mathcal{N}_{M}(t)\delta V_M=N,
\end{equation}
%
and
%
\begin{equation}\label{eq:qGlobRestE}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\epsilon_{n}\right]\delta V_M
    =\sum_{M=1}^{K}\mathcal{E}_M(t)\delta V_M=E. 
\end{equation}
\end{subequations}
%

@@

%============================================================
\subsubsection{Properties of $\mathcal{H}$ for systems in equilibrium}
%============================================================
If the system is in equilibrium (we will show that it will also be valid at $t\to\infty$), the temperature becomes homogeneous throughout
the complete system. Also, the local number of particles, the local
energy, and the Lagrange multipliers do not depend on the cell number and they should be homogeneous,
this is 
%
\begin{subequations}\label{eq:qEqRestrictions}
\begin{equation}
   {\mathcal{N}}_M(t\to\infty)\equiv \bar{\mathcal{N}}=N/K,
\end{equation}
%
%
\begin{equation}
	{\mathcal{E}}_M(t\to\infty)\equiv \bar{\mathcal{E}}=E/K,
\end{equation}
\end{subequations}
%
%
\begin{subequations}\label{eq:qEqAlphaBeta}
\begin{equation}
	\alpha_M=\bar\alpha,\quad\forall M,
\end{equation}
%
and
\begin{equation}
	\beta_M=\bar\beta, \quad\forall M.
\end{equation}
\end{subequations}
%
Substituting Eqs.~(\ref{eq:qEqAlphaBeta}) into Eq.~(\ref{eq:qfMn}) yields the distribution
function of each cell in equilibrium to be
%
\begin{equation}\label{eq:qfneq}
    \bar f_{Mn}=\bar f_n =\frac{1}{\exp\big(-\bar\alpha-\bar\beta \epsilon_n\big)\pm 1},\quad\forall M.
\end{equation}
%
By means of the above equation, we can recover the distribution function and the entropy of a diluted quantum gas in equilibrium as follows. Setting $\bar\alpha=\mu/kT$ and $\bar\beta=-1/kT$,
and substituting them in Eq.~(\ref{eq:qfneq}), it renders the Fermi-Dirac and Bose-Einstein
distribution functions:
%
\begin{equation}\label{eq:qfneqtwo}
    \bar{f}_{n}=\frac{1}{\exp\big(\frac{{\epsilon_n}-\mu}{kT}\big)\pm 1},
\end{equation}
%
and substituting Eq.~(\ref{eq:qfneq}) into the negative of Eq.(\ref{eq:qHdef}), 
the entropy of a quantum ideal gas is obtained
%
\begin{eqnarray}\label{eq:varEntropyDef}
      S&=&\sum_{M=1}^{K}
        \sum_n  \left[
        		\left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
           \ln \left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
         \right]\nonumber \\
      &&\quad\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}
         \left(1 \mp \frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
         \right] \delta V_M.
\end{eqnarray}
%
This quantity is what we will refer to as ``variational entropy'',
and this name reflects the fact that it was obtained \textit{via}
the variational method.

@@

\commr{
An additional consideration is necessary for treating gases composed of fermions,
which must obey the relation $f_{Mn}(t)\leq1$, hence
%
\begin{eqnarray}\label{eq:fermionrestriction}
   1-\bar f_n -\bar f_n g_{nM}\geq0 \ \ \Rightarrow \ \ \frac{1}{\bar f_n}\geq1+g_{nM}.
\end{eqnarray}
%
$\bar f_{n}=1$ is certainly satisfied if the system is at zero temperature. In this state,
all energy levels below and including the Fermi energy are occupied because of
the exclusion principle, therefore the system will
necessarily be homogeneous , and consequently $g_{nM}=0$. In this article, we will
omit this scenario and will only discuss Fermi gases whose temperature is non zero.
}

@@

Succinctly, considering a diluted quantum gas contained in a vessel of volume $V$ (divided into
$K$ small cells), with total energy $E$ and $N$ quantum free particles, which initially
is out of equilibrium ---but in such a manner that the local equilibrium hypothesis is valid---,
the functional
%
\begin{eqnarray}\label{eq:qHfunctTheo}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M,
\end{eqnarray}
%
where $f_{Mn}$ is the $M$-th cell distribution function,
evolves in time such that $d\mathcal{H}/dt\leq0$, and the equality 
condition is attained when the system reaches
the equilibrium state. In Eq.~(\ref{eq:qHfunctTheo}), and for a Fermi (Bose)
gas, $f_{Mn}$ corresponds to the Fermi-Dirac (Bose-Einstein) distrubution function for each cell.
Locally, each cell is in equilibrium, although the complete system may be non-homogeneous,
and is caracterized by the respective $f_{Mn}$, number of particles $\mathcal{N}_M$, energy $\mathcal{E}_M$,
temperature $T_M$, and Legendre Multipliers $\alpha_M$ and $\beta_M$.

>>>>

Consider a dilute gas of $N$ non-interacting quantum particles (either bosons or fermions),
contained by a vessel of volume
$V$, at some temperature $T$, and total energy E. Starting from the Boltzmann definition
of entropy, the quantum $H$ functional is
%
\begin{equation}
   H_T=-\ln G,
\end{equation}
%
where $G$ describes the total number of accessible quantum states of the gas that satisfy the
above conditions \cite{bib:tolman}.
The quantum $H$-theorem can be demostrated as follows.  $G$ can be divided into groups of neighboring states,
$g_k$, and certain occupation numbers, $n_k$, can be associated with each of these
groups. Thus, the above functional takes the form:
%
\begin{equation}\label{eq:quantumh}
    H_T=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, 
\end{equation}
%
where the upper and lower signs are for bosons and fermions, respectively.
Thus the time derivative of Eq.~(\ref{eq:quantumh}) is
%
\begin{equation}\label{eq:quantumdHdt}
\frac{dH_T}{dt}=\sum_{\kappa}\left[\ln n_{\kappa}-\ln\left(g_{\kappa}\pm n_{\kappa}\right)\right]
\frac{dn_{\kappa}}{dt}.
\end{equation}
%

@@

Consider a dilute gas enclosed by a perfectly isolated vessel of volume $V$, with
total energy $E$, and total number of quantum particles $N$, which can be free fermions or
bosons. For our purposes, the volume $V$ is divided into $K$ small cells, each of which
has a constant volume $\delta V_M=V/K$ ($M=1,\,\dots\,,K$), a temperature $T_M$, an energy $\epsilon_M$,
a number of particles $\mathcal{N}_M$, and a distribution function,
$\{f_{Mn}(t)\}$. Hereafter we will use a short-hand notation:
%
\begin{equation}
   f_{Mn}(t)\equiv f(\vec r_M,\epsilon_{n},t),
\end{equation}
%
where $\vec r_M$ is the radius vector pointing at the center of the $M$-th cell.
$f_{Mn}(t)$ represent the number
of particles contained in the $M$-th cell that occupy the energy level $\epsilon_n$ at
time $t$. Since the particles are considered to be free, the energy levels should not
depend on the cell properties, \textit{i.e.} the energy spectrum, $\{\epsilon_n\}$, is the same for all
cells; thus there is no need to label $\epsilon_n$ with an index $M$.

We propose the following functional as an alternative $H$-functional
for quantum non-homogeneous dilute gases:
%
\begin{eqnarray}\label{eq:qHdef}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\bigg.\nonumber \\
    &&\qquad\qquad\bigg.\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M.
\end{eqnarray}
%
Here, the upper and lower sign refer to fermions and bosons respectively.
In addition, when needed, each cell will have an associated local chemical potential,
$\alpha_M$, and a local $H$-functional, which is defined by:
%
\begin{equation}\label{eq:qHMdef}
   \mathcal{H}_M(t)=\sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)
   \pm \Big(1 \mp f_{Mn}(t)\Big) \ln \Big(1 \mp f_{Mn}(t)\Big) \bigg] \de V_M.
\end{equation}
%
Therefore, $\mathcal{N}_M$ and $\mathcal{E}_M$ as functions of time are given by:
%
\begin{subequations}
\begin{equation}
    {\mathcal{N}}_M(t) = \sum_{n}f_{Mn}(t) \delta V_M
\end{equation}
%
and
%
\begin{equation}
\mathcal{E}_M(t) = \sum_{n}f_{Mn}(t)\epsilon_{n} \de V_M,
\end{equation}
\end{subequations}
%
which are, for the whole system, constrained by the microcanonical restrictions
%
\begin{subequations}\label{eq:qGlobRest}
\begin{equation}\label{eq:qGlobRestN}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\right]\delta  V_M
    =\sum_{M=1}^{K}\mathcal{N}_{M}(t)\delta V_M=N,
\end{equation}
%
and
%
\begin{equation}\label{eq:qGlobRestE}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\epsilon_{n}\right]\delta V_M
    =\sum_{M=1}^{K}\mathcal{E}_M(t)\delta V_M=E. 
\end{equation}
\end{subequations}
%

@@

%============================================================
\subsubsection{Properties of $\mathcal{H}$ for systems in equilibrium}
%============================================================
If the system is in equilibrium, the temperature becomes homogeneous throughout
the complete system. Also, the local number of particles, the local
energy, and the Lagrange multipliers do not depend on the cell number and they should be homogeneous,
this is 
%
\begin{subequations}\label{eq:qEqRestrictions}
\begin{equation}
   {\mathcal{N}}_M(t\to\infty)\equiv \bar{\mathcal{N}}=N/K,
\end{equation}
%
%
\begin{equation}
	{\mathcal{E}}_M(t\to\infty)\equiv \bar{\mathcal{E}}=E/K,
\end{equation}
\end{subequations}
%
%
\begin{subequations}\label{eq:qEqAlphaBeta}
\begin{equation}
	\alpha_M=\bar\alpha,\quad\forall M,
\end{equation}
%
and
\begin{equation}
	\beta_M=\bar\beta, \quad\forall M.
\end{equation}
\end{subequations}
%
Substituting Eqs.~(\ref{eq:qEqAlphaBeta}) into Eq.~(\ref{eq:qfMn}) yields the distribution
function of each cell in equilibrium to be
%
\begin{equation}\label{eq:qfneq}
    \bar f_{Mn}=\bar f_n =\frac{1}{\exp\big(-\bar\alpha-\bar\beta \epsilon_n\big)\pm 1},\quad\forall M.
\end{equation}
%
By means of the above equation, we can recover the distribution function and the entropy of a dilute quantum gas in equilibrium as follows. Setting $\bar\alpha=\mu/kT$ and $\bar\beta=-1/kT$,
and substituting them in Eq.~(\ref{eq:qfneq}), it renders the Fermi-Dirac and Bose-Einstein
distribution functions:
%
\begin{equation}\label{eq:qfneqtwo}
    \bar{f}_{n}=\frac{1}{\exp\big(\frac{{\epsilon_n}-\mu}{kT}\big)\pm 1},
\end{equation}
%
and substituting Eq.~(\ref{eq:qfneq}) into the negative of Eq.(\ref{eq:qHdef}), 
the entropy of a quantum ideal gas is obtained
%
\begin{eqnarray}\label{eq:varEntropyDef}
      S&=&\sum_{M=1}^{K}
        \sum_n  \left[
        		\left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
           \ln \left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
         \right]\nonumber \\
      &&\quad\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}
         \left(1 \mp \frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
         \right] \delta V_M.
\end{eqnarray}
%
This quantity is what we will refer to as ``variational entropy'',
and this name reflects the fact that it was obtained \textit{via}
the variational method.

@@

An additional consideration is necessary for treating Fermi gases.
Since, for these systems, $f_{Mn}(t)\leq1$, we have
%
\begin{eqnarray}\label{eq:fermionrestriction}
   1-\bar f_n -\bar f_n g_{nM}\geq0 \ \ \Rightarrow \ \ \frac{1}{\bar f_n}\geq1+g_{nM}.
\end{eqnarray}
%
$\bar f_{n}=1$ is certainly satisfied if the system is at zero temperature. In this state,
all energy levels below and including the Fermi energy are occupied, thus the system will
necessarily be homogeneous , and consequently $g_{nM}=0$. In this article, we will
omit this scenario and will only discuss Fermi gases whose temperature is non zero.

@@

Succinctly, considering a dilute quantum gas contained in a vessel of volume $V$ (divided into
$K$ small cells), with total energy $E$ and $N$ quantum free particles, which initially
is out of equilibrium ---but in such a manner that the local equilibrium hypothesis is valid---,
the functional
%
\begin{eqnarray}\label{eq:qHfunctTheo}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M,
\end{eqnarray}
%
where $f_{Mn}$ is the $M$-th cell distribution function,
evolves in time such that $d\mathcal{H}/dt\leq0$, and the equality 
condition is attained when the system reaches
the equilibrium state. In Eq.~(\ref{eq:qHfunctTheo}), and for a Fermi (Bose)
gas, $f_{Mn}$ corresponds to the Fermi-Dirac (Bose-Einstein) distrubution function for each cell.
Locally, each cell is in equilibrium, although the complete system may be non-homogeneous,
and is caracterized by the respective $f_{Mn}$, number of particles $\mathcal{N}_M$, energy $\mathcal{E}_M$,
temperature $T_M$, and Legendre Multipliers $\alpha_M$ and $\beta_M$.
